Depression and suicidality as costly signals of need

### 2021-04-04 Paper.rmd
Changed adversity section

Alternative depression idea:
- move Firth up before Syme et al. 2016 and tweak slighlty or reword Firth and put behind Syme in the same paragraph
- delete the previous depression stuff
- then follow with this

@@

- emphasize conflict more

Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005; Mazure, 1998; Devries et al. 2011; Ellsberg et al. 2008). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, it's onset tends to quickly follow, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

@@

Of the many types of negative life events associated with depression, those which involve social adversity tend to be more strongly associated with depression (Hammen, 2005; Kendler et al., 1999; Mazure, 1998). Like the relationship between negative events as a whole and depression, this relationship is also seen cross-culturally, with the loss of a loved one, threat of divorce, or abusive relationships all increasing one's chances of becoming depressed (Kendler et al., 1999, 1995 [check if this fits the cross-cultural claim]). Of these events, physical assault has been found to be the strongest predictor of depression in the few studies in which it has been examined (Kendler et al., 1999, 1995) and may have been much more common throughout our evolutionary history than is often appreciated based on rates of spousal abuse in small-scale societies with little or no policing (Levinson, 1989; Stieglitz et al., 2018).

Suicide attempts also tend to co-occur with social adversity. Grief over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016). As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008) and more localized examinations. For example, both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides also reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017).

@@

- add suicide as a pure signal
- merge the hypotheses more

@@

Both depression and suicidal behavior fit this criterion as they are both relatively costly for those not facing adversity than those who are. In the case of depression, the largescale reduction in one's normal routine of fitness enhancing behavior hurts those in need less when one has limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation (Hagen 2003). However, when it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury (Rosenthal, 1993). That given, the previously mentioned logic still applies with increased risk of death being relatively costly for those able to advance their fitness prospects in their day-to-day lives.

Beyond resulting in the cost differential that allows for costly signaling in the first place, the costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial. Importantly, there are likely few forms of social adversity that do not come with conflict. For example, breakups in both romantic and cooperative relationships are often preceded by conflict (Amato & Hohmann-Marriott, 2007), and surely involve conflict when one side wants reconciliation and the other does not. Likewise, the death of a loved one may also be a source of conflict as the survivor may turn to others for more support who might not want to invest more in the survivor (Hagen, 2003). Regardless of the source of conflict, greater conflict should correspond with greater conflicts of interests, something which would favor greater investment in manipulating others through fake signals. In these instances, less costly signals like verbal requests, facial sadness, and crying are likely to be unreliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals in those experiencing need they cannot resolve on their own.

As with any study examining signaling, distinguishing between the information provided by signals vs. cues is a major challenge. A signal differs from a cue in that it is intentionally sent by an organism to alter the behavior or thoughts of other individuals, while a cue is any piece of information an organism uses to alter its behavior or physiological state that was created without the intent to do so (Maynard Smith & Harper, 1995). When it comes to this study, suicide attempts are more likely to take the form of a pure, external signal than depression symptoms. This is because behavioral shifts motivated by low mood for non-signaling benefits would convey information as cues in addition to what might exist for signaling purposes. In other words, the signaling portion of depression may come from greater investment in symptoms than would be optimal for motivating behavior changes alone instead of the creation of entirely new symptoms. As most suicide attempts likely do not have non-signaling functions (see DC for possible exceptions), it is likely much easier to interpret the effect of signaling compared to depression.

In addition to signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support (Hagen 2003). This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

## Responses to depression and suicidal behavior 
For depression and suicide attempts to be worth the costs, substantial benefits in terms of social support are expected, either in the form of temporary but impactful support or long-term changes in how one is treated (Hagen, 2003; Hagen et al., 2008; Syme et al., 2016). This can take the form of reconciliations in failing relationships, increased fairness in cooperative relationships, or an increase of resources being directed towards an individual. Of all the benefits, a reduction in physical abuse may be one of the most important due to the threat of death or severe injury to oneself and one’s family (cite).

>>>>

@@

Likely the most reliable finding regarding depression and suicidality is that most cases are closely tied to major negative life events (Hammen, 2005; Mazure, 1998; Devries et al. 2011; Ellsberg et al. 2008). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events than non-depressed individuals (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In longitudinal studies, the onset of depression tends to quickly follow the negative event (Kendler et al., 1999; search for others) or coincide with periods where conflict is likely to increase prior to it (e.g., depression starting before a divorce rather than after) (Rosenstrom).

@@

Of the types of negative life events associated with depression, those which involve social adversity tend to be more strongly associated with depression (Hammen, 2005; Mazure, 1998). The loss of a loved one, marital problems, bullying (cite), and the existence of abusive relationships are all commonly cited risk factors for depression (Kendler et al., 1999, 1995 [find others]), with  assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression (Kendler et al., 1999, 1995). Other notable risk factors like financial problems, job loss, or severe illness, might seem to lack this social element. However, the ability for others to provide meaningful support to limit the harm caused by such events (sugiyama + others) combined with the ubiquity of conflicts of interests mean that many types of problems will become social even if they did not start that way (Cite you?).

As depression is the leading cause of suicide compared to mental disorders (Hagen 2003 (find more direct citation)), it should not be surprising that suicidality also often co-occurs with social adversity. Grief over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016). As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008) and more localized examinations. For example, both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides also reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017).

@@

@@

Both depression and suicidal behavior fit this criterion as they are both relatively costly for those not facing adversity than those who are. In the case of depression, the largescale reduction in one's normal routine of fitness enhancing behavior hurts those in need less when one has limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation (Hagen 2003). However, when it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury (Rosenthal, 1993). That given, the previously mentioned logic still applies with increased risk of death being relatively more costly for those able to advance their fitness prospects in their day-to-day lives.

Beyond resulting in the cost differential that allows for costly signaling in the first place, the costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial. As previously mentioned, there are likely few forms of social adversity that do not come with conflict. For example, breakups in both romantic and cooperative relationships are often preceded by conflict (Amato & Hohmann-Marriott, 2007; Rosenstrom), and surely involve conflict when one side wants reconciliation and the other does not. Likewise, the death of a loved one may also be a source of conflict as the survivor may turn to others for more support who might not want to invest more in the survivor (Hagen, 2003). Regardless of the source of conflict, greater conflict should correspond with greater conflicts of interests, something which would favor greater investment in manipulating others through fake signals. In these instances, less costly signals like verbal requests, facial sadness, and crying are likely to be unreliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals in those experiencing need they cannot resolve on their own.

As with any study examining signaling, distinguishing between the information provided by signals versus cues is a major challenge. A signal differs from a cue in that it is intentionally sent by an organism to alter the behavior or thoughts of other individuals, while a cue is any piece of information an organism uses to alter its behavior or physiological state that was created without the intent to do so (Maynard Smith & Harper, 1995). Regarding the signals of interest, suicide attempts are more likely to take the form of a pure, external signal than depression symptoms. This is because behavioral shifts motivated by low mood for non-signaling benefits would convey information as cues in addition to what might exist for signaling purposes. In other words, the signaling portion of depression may come from the upregulation of symptoms beyond what would be optimal for motivating behavior changes alone instead of the creation of entirely new symptoms. As most suicide attempts likely do not have non-signaling functions (see DC for a possible exception), this is less likely to be a challenge with suicide attempts.

Beyond signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support (Hagen 2003). This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

## Responses to depression and suicidal behavior

For depression and suicide attempts to be worth the costs involved with signaling, substantial benefits in terms of social support are expected, either in the form of temporary but impactful support or long-term changes in how one is treated (Hagen, 2003; Hagen et al., 2008; Syme et al., 2016). This can take the form of reconciliations in failing relationships, increased fairness in cooperative relationships, or an increase of resources being directed towards an individual. Of all the benefits, a reduction in physical abuse may be one of the most important due to the threat of death or severe injury to oneself and one’s family (cite).
